---
title: "NHANES download"
format: html
date: "today"
---

Download the 2013-2018 NHANES public datasets (demographics, healthcare
utilization, health insurance, blood-pressure questionnaire, blood-pressure
exam, household, income), left-join on `SEQN`, and write `nhanes.csv`. This is
the raw data layer for both `NHANES_multivariate_copula.qmd` (which fits the
copula + samples the simulated cohort) and `NHANES_out_of_system_BP.qmd` (which
fits the out-of-system BP regressions).

```{r}
#| label: setup
library(haven)
library(here)
library(purrr)
library(dplyr)
library(readr)

data_dir <- here::here("data_and_models/nhanes_data_cohorts")
if (!dir.exists(data_dir)) {
  dir.create(data_dir)
}
```

```{r}
#| label: helpers
download_and_read_xpt <- function(url) {
  .tf <- tempfile()
  download.file(url, .tf, mode = "wb")

  read_xpt(.tf)
}

load_nhanes_and_select <- function(urls, ...) {
  map(urls, download_and_read_xpt) |>
    list_rbind() |>
    select(...)
}
```

## Demographics 2013-2018

```{r}
#| label: load_demographics
demographic_urls <- c(
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/DEMO_J.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/DEMO_I.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/DEMO_H.xpt"
)

demo <- load_nhanes_and_select(
  demographic_urls,
  SEQN, RIAGENDR, RIDAGEYR, SDMVSTRA,
  SDMVPSU, WTMEC2YR, RIDRETH3, INDFMPIR, DMDEDUC2, DMDHHSIZ, DMDMARTL,
  DMDHHSZA, DMDHHSZB
)
```

## Healthcare utilization questions 2013-2018

```{r}
#| label: load_huq
huq_urls <- c(
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/HUQ_J.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/HUQ_I.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/HUQ_H.xpt"
)

huq <- load_nhanes_and_select(
  huq_urls,
  SEQN, HUQ030, HUQ041, HUQ051, HUQ061
)
```

## Health insurance 2013-2018

```{r}
#| label: load_hiq
hiq_urls <- c(
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/HIQ_J.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/HIQ_I.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/HIQ_H.xpt"
)

hiq <- load_nhanes_and_select(
  hiq_urls,
  SEQN, HIQ011
)
```

## Blood pressure questions 2013-2018

```{r}
#| label: load_bpq
bpq_urls <- c(
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/BPQ_J.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/BPQ_I.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/BPQ_H.xpt"
)

bpq <- load_nhanes_and_select(
  bpq_urls,
  SEQN, BPQ020, BPQ030, BPQ040A, BPQ050A
)
```

# Blood pressure lab values 2013-2018

```{r}
#| label: load_bpx
bpx_urls <- c(
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/BPX_J.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/BPX_I.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/BPX_H.xpt"
)

bpx <- load_nhanes_and_select(
  bpx_urls,
  SEQN, BPXSY1, BPXDI1
)
```

# Household characteristics 2013-2018

```{r}
#| label: load_hoq
hoq_urls <- c(
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/HOQ_J.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/HOQ_I.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/HOQ_H.xpt"
)

hoq <- load_nhanes_and_select(
  hoq_urls,
  SEQN, HOD050, HOQ065
)
```

# Income 2013-2018

```{r}
#| label: load_inq
inq_urls <- c(
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/INQ_J.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/INQ_I.xpt",
  "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/INQ_H.xpt"
)

inq <- load_nhanes_and_select(
  inq_urls,
  SEQN, INQ020, INQ012
)
```

```{r}
#| label: join_and_write
nhanes_data <- list(demo, huq, hiq, bpq, bpx, hoq, inq) |>
  reduce(left_join, by = "SEQN")

write_csv(nhanes_data, file.path(data_dir, "nhanes.csv"))
```
