---
title: "NHANES multivariate copula"
format: html
date: "today"
---

```{r}
#| label: setup
library(tidyverse)
library(survey)
library(copula)
library(openxlsx)
library(jtools)
library(here)
library(jsonlite)
library(purrr)
library(dplyr)
library(readr)
# also using the `metan::mantel_test()`
# but loading metan masks a lot of functions

data_dir <- here::here("data_and_models/nhanes_data_cohorts")
set.seed(123)
```

Read the joined NHANES dataset produced by `NHANES_download.qmd`.

```{r}
#| label: load_nhanes
nhanes_data <- read_csv(file.path(data_dir, "nhanes.csv"))
```

```{r}
# adjust weights and filter data
nhanes <- nhanes_data |>
  mutate(
    WTMEC6YR = WTMEC2YR / 3,
    female = if_else(RIAGENDR == 2, 1, 0),
    black = if_else(RIDRETH3 == 4, 1, 0),
    white = if_else(RIDRETH3 == 3, 1, 0),
    place = if_else((HUQ030 != 2) & (HUQ030 != 7), 1, 0),
    insurance = if_else(HIQ011 == 1, 1, 0),
    treated = if_else(BPQ050A == 1, 1, 0),
    poor = if_else(INDFMPIR < 1, 1, 0),
    education_L12 = ifelse(DMDEDUC2 == 1 | DMDEDUC2 == 2, 1, 0),
    rented = ifelse(HOQ065 == 2, 1, 0),
    notemployed = ifelse((INQ020 == 2 & INQ012 == 2) & RIDAGEYR < 65, 1, 0),
    crowded = ifelse(DMDHHSIZ - HOD050 > 0, 1, 0),
    singlefamily = ifelse((DMDHHSZA > 0 | DMDHHSZB > 0) & (DMDMARTL == 2 |
      DMDMARTL == 3 |
      DMDMARTL == 4 |
      DMDMARTL == 5), 1, 0)
  ) |>
  replace_na(list(treated = 0)) |>
  filter(!is.na(RIDAGEYR), RIDAGEYR > 18) |>
  # self-reported number of times receiving healthcare over the past year
  # we are assuming equal chance of each number when reporting a range
  mutate(visits = case_when(
    HUQ051 == 0 ~ 0,
    HUQ051 == 1 ~ 1,
    HUQ051 == 2 ~ sample(c(2, 3), n(), replace = TRUE), # self-reported 2 to 3
    HUQ051 == 3 ~ sample(c(4, 5), n(), replace = TRUE), # self-reported 4 to 5
    HUQ051 == 4 ~ sample(c(6, 7), n(), replace = TRUE), # self-reported 6 to 7
    HUQ051 == 5 ~ sample(c(8, 9), n(), replace = TRUE), # self-reported 8 to 9
    HUQ051 == 6 ~ sample(c(10, 11, 12), n(), replace = TRUE), # self-reported 10 to 12
    HUQ051 == 7 ~ sample(c(13, 14, 15), n(), replace = TRUE), # self-reported 13 to 15
    HUQ051 == 8 ~ sample(c(16, 17, 18), n(), replace = TRUE), # self-reported 16 or more
    HUQ051 == 77 ~ 0, # refused
    HUQ051 == 99 ~ 0 # don't know
  )) |>
  mutate(across(c(BPXSY1, BPXDI1), ~ na_if(.x, 0))) |>
  filter(treated == 0, (black == 1 | white == 1))

nhanes_svy <- svydesign(
  data = nhanes,
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC6YR,
  nest = TRUE
)
```

```{r}
#not going to also filter by individuals with or without treatment
nhanes2 <- nhanes_data |>
  mutate(
    WTMEC6YR = WTMEC2YR / 3,
    female = if_else(RIAGENDR == 2, 1, 0),
    black = if_else(RIDRETH3 == 4, 1, 0),
    white = if_else(RIDRETH3 == 3, 1, 0),
    place = if_else((HUQ030 != 2) & (HUQ030 != 7), 1, 0),
    insurance = if_else(HIQ011 == 1, 1, 0),
    treated = if_else(BPQ050A == 1, 1, 0),
    poor = if_else(INDFMPIR < 1, 1, 0),
    education_L12 = ifelse(DMDEDUC2 == 1 | DMDEDUC2 == 2, 1, 0),
    rented = ifelse(HOQ065 == 2, 1, 0),
    notemployed = ifelse((INQ020 == 2 & INQ012 == 2) & RIDAGEYR < 65, 1, 0),
    crowded = ifelse(DMDHHSIZ - HOD050 > 0, 1, 0),
    singlefamily = ifelse((DMDHHSZA > 0 | DMDHHSZB > 0) & (DMDMARTL == 2 |
      DMDMARTL == 3 |
      DMDMARTL == 4 |
      DMDMARTL == 5), 1, 0)
  ) |>
  replace_na(list(treated = 0)) |>
  filter(!is.na(RIDAGEYR), RIDAGEYR > 18) |>
  mutate(visits = case_when(
    HUQ051 == 0 ~ 0,
    HUQ051 == 1 ~ 1,
    HUQ051 == 2 ~ sample(c(2, 3), n(), replace = TRUE), # self-reported 2 to 3
    HUQ051 == 3 ~ sample(c(4, 5), n(), replace = TRUE), # self-reported 4 to 5
    HUQ051 == 4 ~ sample(c(6, 7), n(), replace = TRUE), # self-reported 6 to 7
    HUQ051 == 5 ~ sample(c(8, 9), n(), replace = TRUE), # self-reported 8 to 9
    HUQ051 == 6 ~ sample(c(10, 11, 12), n(), replace = TRUE), # self-reported 10 to 12
    HUQ051 == 7 ~ sample(c(13, 14, 15), n(), replace = TRUE), # self-reported 13 to 15
    HUQ051 == 8 ~ sample(c(16, 17, 18), n(), replace = TRUE), # self-reported 16 or more
    HUQ051 == 77 ~ 0, # refused
    HUQ051 == 99 ~ 0 # don't know
  )) |>
  mutate(across(c(BPXSY1, BPXDI1), ~ na_if(.x, 0))) |>
  filter(black == 1 | white == 1)

nhanes_svy2 <- svydesign(
  data = nhanes2,
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC6YR,
  nest = TRUE
)
# Define 5-year age bins: 40-44, 45-49, 50-54, ...
max_age <- max(nhanes$RIDAGEYR, na.rm = TRUE)
age_breaks <- seq(40, floor(max_age / 5) * 5 + 5, by = 5)
age_labels <- paste0(age_breaks[-length(age_breaks)], "-", age_breaks[-1] - 1)

# Restrict to age >= 40 and add age-group variable inside the survey design
nhanes_svy_40plus <- subset(nhanes_svy2, RIDAGEYR >= 40)

nhanes_svy_40plus <- update(
  nhanes_svy_40plus,
  age_group = cut(
    RIDAGEYR,
    breaks = age_breaks,
    labels = age_labels,
    right = FALSE,
    include.lowest = TRUE
  )
)

# Survey-weighted mean SBP and DBP by age group
bp_means_by_age <- svyby(
  ~BPXSY1 + BPXDI1,
  ~age_group,
  design = nhanes_svy_40plus,
  FUN = svymean,
  na.rm = TRUE,
  vartype = "se"
) |>
  as.data.frame() |>
  rename(
    mean_sbp = BPXSY1,
    mean_dbp = BPXDI1,
    se_sbp = se.BPXSY1,
    se_dbp = se.BPXDI1
  )

print(bp_means_by_age)

write.csv(bp_means_by_age, file.path(here::here("data_and_models"), "nhanes_inputs/bp_means_by_age_5yr.csv"))
```


```{r}
survey_corr <- svycor(
  ~ RIDAGEYR + female + black + place + insurance + INDFMPIR + BPXSY1 + BPXDI1 +
    poor + education_L12 + rented + notemployed + crowded + singlefamily + visits,
  na.rm = TRUE,
  design = nhanes_svy
)

survey_corr <- survey_corr$cors
```

```{r}
female_prop <- svymean(~female, nhanes_svy)[[1]]
black_prop <- svymean(~black, nhanes_svy)[[1]]
white_prop <- svymean(~white, nhanes_svy)[[1]]
place_prop <- svymean(~place, nhanes_svy)[[1]]
insurance_prop <- svymean(~insurance, nhanes_svy)[[1]]
poor_prop <- svymean(~poor, nhanes_svy, na.rm = TRUE)[[1]]
education_L12_prop <- svymean(~education_L12, nhanes_svy, na.rm = TRUE)[[1]]
rented_prop <- svymean(~rented, nhanes_svy, na.rm = TRUE)[[1]]
notemployed_prop <- svymean(~notemployed, nhanes_svy, na.rm = TRUE)[[1]]
crowded_prop <- svymean(~crowded, nhanes_svy, na.rm = TRUE)[[1]]
singlefamily_prop <- svymean(~singlefamily, nhanes_svy, na.rm = TRUE)[[1]]

age_min <- 19
age_max <- 80

BPXSY1_mean <- svymean(~BPXSY1, nhanes_svy, na.rm = TRUE)[[1]]
BPXSY1_var <- svyvar(~BPXSY1, nhanes_svy, na.rm = TRUE)[[1]]
BPXDI1_mean <- svymean(~BPXDI1, nhanes_svy, na.rm = TRUE)[[1]]
BPXDI1_var <- svyvar(~BPXDI1, nhanes_svy, na.rm = TRUE)[[1]]
RIDAGEYR_mean <- svymean(~RIDAGEYR, nhanes_svy)[[1]]
RIDAGEYR_var <- svyvar(~RIDAGEYR, nhanes_svy)[[1]]
INDFMPIR_mean <- svymean(~INDFMPIR, nhanes_svy, na.rm = TRUE)[[1]]
INDFMPIR_var <- svyvar(~INDFMPIR, nhanes_svy, na.rm = TRUE)[[1]]
visits_mean <- svymean(~visits, nhanes_svy, na.rm = TRUE)[[1]]
visits_var <- svyvar(~visits, nhanes_svy, na.rm = TRUE)[[1]]
```

## Distribution for family income to poverty ratio

```{r}
svyhist(~INDFMPIR, nhanes_svy)
# shape and rate determined from trial and error to match the distribution above
poverty_ratio_shape <- 3
poverty_ratio_rate <- 0.75
betas <- rgamma(1000, shape = poverty_ratio_shape, rate = poverty_ratio_rate)
betas <- if_else(betas > 5, 5, betas)
hist(betas, freq = FALSE)
```

## Distribution for systolic and diastolic blood pressure

```{r}
svyhist(~BPXDI1, nhanes_svy)
hist(rnorm(1000, mean = BPXDI1_mean, sd = sqrt(BPXDI1_var)))

svyhist(~BPXSY1, nhanes_svy)
hist(rnorm(1000, mean = BPXSY1_mean, sd = sqrt(BPXSY1_var)))
```
## Distribution of visits in the past year
```{r}
svyhist(~visits, nhanes_svy)
visits_size <- 5
# size and mu determined from trial and error to match the distribution above
betas <- rnbinom(1000, size = visits_size, mu = visits_mean)
var(betas)
hist(betas, freq = FALSE)
```
I slightly strengthened the correlations, as the simulated cohort output 
correlations were not matching the NHANES correlations
```{r}
orig_correlations <- survey_corr[lower.tri(survey_corr)]
correlations <- survey_corr[lower.tri(survey_corr)]
change_df <- read_csv(file.path(here::here("data_and_models"), "nhanes_inputs/change_corr.csv"), show_col_types = FALSE)
change_list <- change_df$index
change_add <- change_df$addition

# Find the indices of correlations that are in change_list
indices <- match(change_list, seq_along(correlations))

# Directly add the corresponding values from change_add
correlations[indices] <- correlations[indices] + change_add
```

Fit multivariate normal copula to marginal distributions 
and survey correlation matrix

```{r}
myCop <- normalCopula(param = correlations, dim = 15, dispstr = "un")
myMvd <- mvdc(
  copula = myCop,
  margins = c(
    "unif", "binom", "binom",
    "binom", "binom", "gamma",
    "norm", "norm", "binom", "binom",
    "binom", "binom", "binom", "binom",
    "nbinom"
  ),
  paramMargins = list(
    # age
    list(min = age_min, max = age_max),
    # female
    list(size = 1, prob = female_prop),
    # black
    list(size = 1, prob = black_prop),
    # place
    list(size = 1, prob = place_prop),
    # insurance
    list(size = 1, prob = insurance_prop),
    # Poverty ratio
    list(shape = poverty_ratio_shape, rate = poverty_ratio_rate),
    # systolic
    list(mean = BPXSY1_mean, sd = sqrt(BPXSY1_var)),
    # diastolic
    list(mean = BPXDI1_mean, sd = sqrt(BPXDI1_var)),
    # poor
    list(size = 1, prob = poor_prop),
    # education_L12
    list(size = 1, prob = education_L12_prop),
    # rented
    list(size = 1, prob = rented_prop),
    # notemployed
    list(size = 1, prob = notemployed_prop),
    # crowded
    list(size = 1, prob = crowded_prop),
    # singlefamily
    list(size = 1, prob = singlefamily_prop),
    # visits
    list(size = 5, mu = visits_mean)
  )
)

simulated_data <- rMvdc(1000000, myMvd)
colnames(simulated_data) <- c(
  "age", "female", "black", "place",
  "insurance", "povertyratio",
  "systolicbp", "diastolicbp", "poor",
  "educationL12", "rented", "notemployed",
  "crowded", "singlefamily", "visits"
)

simulated_df <- as_tibble(simulated_data) |>
  mutate(povertyratio = if_else(povertyratio > 5, 5, povertyratio))
summary(simulated_df)
```


```{r}
if (!dir.exists(file.path(here::here("data_and_models"), "nhanes_inputs/copula_inputs/"))) {
  dir.create(file.path(here::here("data_and_models"), "nhanes_inputs/copula_inputs/"))
}
if (!dir.exists(file.path(here::here("data_and_models"), "nhanes_inputs/copula_outputs/"))) {
  dir.create(file.path(here::here("data_and_models"), "nhanes_inputs/copula_outputs/"))
}
```

Save the parameters to be read into quarto document
```{r}
variables <- list(
  female = list(data = female_prop, file = "female_prop.json"),
  black = list(data = black_prop, file = "black_prop.json"),
  place = list(data = place_prop, file = "place_prop.json"),
  insurance = list(data = insurance_prop, file = "insurance_prop.json"),
  poor = list(data = poor_prop, file = "poor_prop.json"),
  education_L12 = list(data = education_L12_prop, file = "education_L12_prop.json"),
  rented = list(data = rented_prop, file = "rented_prop.json"),
  notemployed = list(data = notemployed_prop, file = "notemployed_prop.json"),
  crowded = list(data = crowded_prop, file = "crowded_prop.json"),
  singlefamily = list(data = singlefamily_prop, file = "singlefamily_prop.json"),
  age_min = list(data = age_min, file = "age_min.json"),
  age_max = list(data = age_max, file = "age_max.json"),
  poverty_ratio_shape = list(data = poverty_ratio_shape, file = "poverty_ratio_shape.json"),
  poverty_ratio_rate = list(data = poverty_ratio_rate, file = "poverty_ratio_rate.json"),
  RIDAGEYR_mean = list(data = RIDAGEYR_mean, file = "age_mean.json"),
  age_std = list(data = sqrt(RIDAGEYR_var), file = "age_std.json"),
  INDFMPIR_mean = list(data = INDFMPIR_mean, file = "poverty_ratio_mean.json"),
  poverty_ratio_std = list(data = sqrt(INDFMPIR_var), file = "poverty_ratio_std.json"),
  BPXSY1_mean = list(data = BPXSY1_mean, file = "systolic_mean.json"),
  systolic_std = list(data = sqrt(BPXSY1_var), file = "systolic_std.json"),
  BPXDI1_mean = list(data = BPXDI1_mean, file = "diastolic_mean.json"),
  diastolic_std = list(data = sqrt(BPXDI1_var), file = "diastolic_std.json"),
  visits_mean = list(data = visits_mean, file = "visits_mean.json"),
  visits_std = list(data = sqrt(visits_var), file = "visits_std.json"),
  visits_size = list(data = visits_size, file = "visits_size.json")
)

write_variable_json <- function(variable, base_path) {
  write_json(variable$data, file.path(base_path, variable$file))
}

base_path <- here::here("data_and_models", "nhanes_inputs", "copula_inputs")

walk(variables, write_variable_json, base_path = base_path)
```

Save the simulated cohort summary statistics
```{r}
# Calculate means and standard deviations
stats <- simulated_df |>
  summarise(across(everything(), list(mean = mean, std = sd))) |>
  pivot_longer(cols = everything(), names_to = c("variable", ".value"), names_sep = "_") |>
  (\(df) split(df, df$variable))() |>
  lapply(\(x) list(mean = x$mean, std = x$std))

# Convert to JSON
stats_json <- toJSON(stats, pretty = TRUE)

write(stats_json, file.path(here::here("data_and_models"), "nhanes_inputs/copula_outputs/summary_stats.json"))
```

Mantel test to test similarity between original NHANES and simulated data correlation matrices
```{r}
metan::mantel_test(survey_corr, cor(simulated_data))
```
Output correlation matrices from NHANES and simulated cohort
```{r}
## correlations from NHANES data
survey_corr_long <- survey_corr |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    -var1,
    names_to = "var2",
    values_to = "correlation"
  )
rename_var <- function(x) {
  case_match(
    as.character(x),
    "RIDAGEYR" ~ "age",
    "INDFMPIR" ~ "povertyratio",
    "BPXSY1" ~ "systolicbp",
    "BPXDI1" ~ "diastolicbp",
    "education_L12" ~ "educationL12",
    .default = as.character(x)
  )
}

survey_corr_long <- survey_corr_long |>
  mutate(across(c(var1, var2), rename_var))

# Correlations from simulated copula data
cor_long <- cor(simulated_data) |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(
    -var1,
    names_to = "var2",
    values_to = "correlation"
  )

combined_correlations <- survey_corr_long |>
  left_join(cor_long, by = c("var1", "var2"))

combined_correlations <- combined_correlations |>
  filter(var1 != var2) |>
  distinct(correlation.x, correlation.y, .keep_all = TRUE)

combined_correlations
write.csv(combined_correlations, file.path(here::here("data_and_models"), "nhanes_inputs/copula_outputs/correlations.csv"))
```

Sample cohort (50,000 NHB and 50,000 NHW)
```{r}
df_around_40 <- simulated_df |>
  filter(between(age, 35, 45))

sample_by_race <- function(.df, n) {
  .df |>
    group_by(black) |>
    sample_n(size = n, replace = TRUE) |>
    ungroup()
}

combined_df <- sample_by_race(df_around_40, 50000)

combined_df |>
  group_by(black) |>
  group_map(summary)

write_csv(
  combined_df,
  file.path(data_dir, "NHANES_final_cohort.csv")
)
```


```{r}
# Calculate means and standard deviations
cohort_stats <- combined_df |>
  group_by(black) |>
  summarise(across(everything(), list(mean = mean, std = sd))) |>
  pivot_longer(cols = everything(), names_to = c("variable", ".value"), names_sep = "_") |>
  (\(df) split(df, df$variable))() |>
  lapply(\(x) list(mean = x$mean, std = x$std))

# Convert to JSON
cohort_stats_json <- toJSON(cohort_stats, pretty = TRUE)

write(cohort_stats_json, file.path(here::here("data_and_models"), "nhanes_inputs/copula_outputs/cohort_summary_stats_final.json"))
```

Output calibration targets directly from NHANES
```{r}
nhanes <- nhanes |>
  mutate(
    agegrp = cut(RIDAGEYR, c(19, 24, 34, 44, 54, 64, 74, 84), labels = c("19-24", "25-34", "35-44", "45-54", "55-64", "65-74", "75+")),
    noplace = if_else(HUQ030 %in% c(2, 7, 9), 1, 0),
  )
nhanes_svy <- svydesign(
  data = nhanes,
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC6YR,
  nest = TRUE
)

noplace_df <- svyby(~noplace, ~ agegrp + insurance + black, design = nhanes_svy, svymean)
write_csv(
  noplace_df,
  file.path(here::here("data_and_models"), "nhanes_inputs/copula_outputs/nhanes_place_calibration_targets.csv")
)
```


